---
title: "Review of SVP calibration for the well-log data"
author: "svpChange simulation audit"
date: "15 September 2026"
output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 10pt
geometry: margin=0.8in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
review <- readRDS("well_log_review.rds")
meta <- review$metadata
observed <- review$observed$table
calibration <- review$calibration$summary
bootstrap_summary <- review$bootstrap_summary
```

# Recommendation

For this data, the preferred calibrated configuration is

```{r recommended-call, results='asis'}
cat(
  "```r\n",
  "SVP(y, gamma = ", round(meta$selected_gamma, 3), ",\n",
  "    test = \"MedianMoodCost\", subtests = \"right\", ",
  "cost = \"gaussian\")\n",
  "```",
  sep = ""
)
```

The threshold is the empirical 0.99 quantile of the maximum Median--Mood
statistic over every growing prefix under a dependence-preserving residual
block bootstrap. This is exactly the one-segment criterion for `right`; it is
not selected to maximize agreement with RFPOP. The recommendation changes the
current application from `both` to `right`, removes the oracle segment count,
and accounts for serial dependence.

# Existing analysis and literature benchmark

The current application contains two incompatible oracle segment counts: 30
for Wilcoxon and 10 for Median--Mood. Its Mood threshold also uses a
$\chi^2_{K}$ quantile, although the Pearson statistic has one degree of
freedom. Neither construction accounts for the substantial serial dependence
remaining in the well-log data.

The external benchmark is the existing RFPOP fit with biweight loss,
`lambda = 70`, and `lthreshold = 2`. This reproduces the application of
[Fearnhead and Rigaill (2019)](https://doi.org/10.1080/01621459.2017.1385466),
who use a threshold of twice the estimated noise scale and a penalty of 70
estimated noise variances. It gives the following internal changepoints:

```{r reference}
review$reference
```

RFPOP is treated as an external benchmark, not ground truth. In particular,
using a first-difference MAD rather than the package application's global MAD
adds a short segment near positions 1034--1069. The 19-change analysis in
[Wilms, Killick, and Matteson (2022)](https://doi.org/10.1080/10618600.2021.2000873)
is not a compatible gold standard: it analyzes a 1,000-observation subset with
Gaussian PELT and emphasizes the instability of that segmentation.

```{r rfpop-sensitivity}
knitr::kable(
  review$rfpop_scale_sensitivity[, c(
    "scale_method", "scale", "n_changepoints"
  )],
  digits = 2, booktabs = TRUE,
  caption = "Sensitivity of the RFPOP benchmark to noise-scale estimation."
)
```

The global-MAD fit gives changes at
`r gsub(",", ", ", review$rfpop_scale_sensitivity$changepoints[review$rfpop_scale_sensitivity$scale_method == "package_global_mad"])`.
The first-difference-MAD fit gives changes at
`r gsub(",", ", ", review$rfpop_scale_sensitivity$changepoints[review$rfpop_scale_sensitivity$scale_method == "first_difference_mad"])`.

# Null calibration

We subtract the median in each RFPOP segment and resample residual blocks that
do not cross a reference boundary. The block length is the first lag followed
by five residual autocorrelations inside $\pm1.96/\sqrt{n}$; this gives
`r meta$selected_block_length`. Calibration and independent validation use
`r format(meta$calibration_reps, big.mark = ",")` series each, with block
lengths 20 and 64 included as sensitivity checks. A null series is counted as
correct only when the complete `right` run returns one segment.

```{r calibration-table}
display_calibration <- calibration
display_calibration$gamma <- round(display_calibration$gamma, 3)
for (column in c("calibration_recovery", "validation_recovery",
                 "validation_lower", "validation_upper")) {
  display_calibration[[column]] <- round(display_calibration[[column]], 4)
}
knitr::kable(
  display_calibration[, c(
    "test", "block_length", "gamma", "calibration_recovery",
    "validation_recovery", "validation_lower", "validation_upper"
  )],
  booktabs = TRUE,
  caption = "Block-bootstrap null calibration and independent validation."
)
```

```{r null-figure, out.width="95%", fig.align="center"}
knitr::include_graphics("plots/02_null_calibration.pdf")
```

The Wilcoxon scan is strongly inflated by the residual dependence. Its
calibrated threshold is therefore large and retains only the most prominent
transitions. The oracle-$K$ Wilcoxon scaling used in the power experiment is
not suitable for this real-data application.

The same threshold cannot be described as a `both` calibration. In that mode,
an invalid interior candidate can remove the one-segment candidate even when
all full-prefix tests pass. The direct stress check below therefore reports
the actual `both` one-segment rate at the `right` threshold.

```{r both-null-table}
knitr::kable(
  review$both_null_stress[, c(
    "subtests", "gamma", "one_segment_recovery", "lower", "upper", "reps"
  )],
  digits = 3, booktabs = TRUE,
  caption = paste0(
    "Direct null stress check for both over ", meta$both_null_reps,
    " dependent bootstrap series."
  )
)
```

# Observed segmentations

All comparisons below remove the terminal endpoint and use one-to-one matching
within 10 observations. These scores measure agreement with RFPOP rather than
accuracy against known geological boundaries.

```{r observed-table}
display_observed <- observed[, c(
  "method", "n_changepoints", "matched", "precision", "recall", "F1"
)]
display_observed$method <- sub("Calibrated SVP ", "SVP ",
                               display_observed$method, fixed = TRUE)
knitr::kable(
  display_observed, digits = 3, booktabs = TRUE,
  caption = "Observed-data agreement with the external RFPOP benchmark."
)
```

```{r segmentation-figure, out.width="100%", fig.align="center"}
knitr::include_graphics("plots/01_segmentations.pdf")
```

The calibrated `right` fit matches seven of the ten RFPOP boundaries and adds
one SVP-only candidate near 3744. It merges several short neighboring segments
in the 2400--2600 cluster; this is the cost of controlling false segmentation
under the strongly dependent null. At the same threshold, `both` does not
improve agreement because some clustered boundaries shift outside the matching
tolerance. The legacy Mood and Wilcoxon settings over-segment substantially.

# Mode and threshold sensitivity

The signal bootstrap adds independently resampled residual blocks to the
piecewise-median RFPOP signal. It is therefore a model-assisted comparison of
the two SVP modes, not independent geological validation.

```{r bootstrap-table}
display_bootstrap <- bootstrap_summary[, c(
  "mode", "mean_precision", "mean_recall", "mean_F1",
  "mean_changepoints", "reps"
)]
names(display_bootstrap) <- c(
  "mode", "precision", "recall", "F1", "mean CPs", "reps"
)
knitr::kable(
  display_bootstrap, digits = 3, booktabs = TRUE,
  caption = paste0(
    "Mode comparison over ", meta$power_reps,
    " signal-plus-residual bootstrap series."
  )
)
```

Across these replicates, `right` has mean F1
`r sprintf("%.3f", bootstrap_summary$mean_F1[bootstrap_summary$mode == "right"])`,
compared with
`r sprintf("%.3f", bootstrap_summary$mean_F1[bootstrap_summary$mode == "both"])`
for `both` at the same threshold.

```{r bootstrap-figure, out.width="95%", fig.align="center"}
knitr::include_graphics("plots/03_mode_bootstrap.pdf")
```

```{r frequency-figure, out.width="98%", fig.align="center"}
knitr::include_graphics("plots/04_selection_frequency.pdf")
```

```{r threshold-figure, out.width="95%", fig.align="center"}
knitr::include_graphics("plots/05_threshold_sensitivity.pdf")
```

# Interpretation and limitations

The calibrated Median--Mood `right` configuration is the defensible replacement
for the current application settings: it removes the unknown oracle segment
count and controls the whole-series null false-segmentation probability while
retaining the main RFPOP transitions. The `both` mode can look closer to RFPOP
at a lower threshold, but the null experiment shows that this comes from
uncontrolled interior-subsegment multiplicity rather than better calibration.

This remains an empirical recommendation. RFPOP is not geological ground
truth, and the bootstrap null is conditional on its fitted segmentation. The
`right` result is also direction-dependent; the reverse-series comparison is
reported in the CSV output. Exact pruning equivalence for the Median--Mood
validity rule has not been proved. Finally, the Gaussian segment cost is only
SVP's tie-breaker among partitions with the same number of segments; a robust
tie-breaking cost is a separate possible extension.

# Reproducibility

The analysis used `svpChange` `r meta$svpChange_version` from
`r meta$svpChange_library`, `robseg` `r meta$robseg_version`, data length
`r meta$n`, and data SHA-256 fingerprint `r meta$data_sha256`. Calibration,
validation, and signal-bootstrap seeds are `r meta$calibration_seed`,
`r meta$validation_seed`, and `r meta$power_seed`, respectively.
