---
title: "Corrected Section 5.3 AR(1) power study"
author: "svpChange"
date: "2026-09-15"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
project_root <- normalizePath(
  file.path(dirname(knitr::current_input(dir = TRUE)), "..", "..")
)
knitr::opts_knit$set(root.dir = project_root)
setwd(project_root)
source(file.path("simulations", "power_ar1", "run_power.R"))
```

# Study model

The experiment follows the paper's Section 5.3 design:

\[
Y_t = \mu_t + E_t, \qquad E_t = \rho E_{t-1} + \eta_t,
\qquad \rho=0.8.
\]

The noise is generated by `ts_generator(type = "gaussAR1")` with
`sd_noise = sqrt(1-rho^2)`. Its stationary marginal variance is therefore one,
while the innovation variance supplied to DeCAFS and AR1Focus is
`sigma2 = 1-rho^2 = 0.36`. The first residual is drawn from the stationary
distribution.

# Signal design and metrics

The four signals are `none`, `up`, `updown`, and `rand1`, using the shared
`generate_power_signal()` implementation. We use `n = 600`, 100 replications,
and jump sizes `seq(1, 5, 0.2)`, giving 8,400 simulated series.

The mandatory terminal endpoint `n` is removed before scoring. Internal
changepoints are matched one-to-one within the Section 5.3 tolerance of five
observations. The study records precision, recall, F1, probability of the
correct number of changepoints, signal MSE, fitted segment count, and maximum
localization error when the number of changes is correct.

```{r design-check}
stopifnot(
  identical(AR1_N, 600L),
  identical(AR1_JUMP_SIZES, seq(1, 5, 0.2)),
  length(AR1_JUMP_SIZES) == 21L,
  AR1_REPS == 100L,
  AR1_TOLERANCE == 5L
)
```

# DeCAFS version

DeCAFS 3.3.6 is required, and `calibrate_ar1.R` stops against anything older.
The exact installed version is recorded in the calibration output.
Equation (4) of the DeCAFS paper, and the package's own README, put the
stationary AR(1) precision \((1-\phi^2)/\sigma_\nu^2\) on
\((y_1-\mu_1)^2\); builds before 3.3.6 initialised the dynamic programme with
\(1/(\sigma_\nu^2(1-\phi^2))\) instead. At `rho = 0.8` that over-weights the
first observation by \((1-\rho^2)^{-2} = 7.7\), pinning \(\mu_1\) to
\(y_1\) so that a change in the first few observations pays for itself. Under
the null at `beta = 2 log(n)`, 37% of series then carried a false change and
about half of those lay in the first 20 points. The calibration below removes
them by raising the penalty, which is why the older build selected
`7.75 log(n)` and lost power against every alternative. With the fix the
selected constant is `2.75 log(n)`, next to the package default `2 log(n)`, and
DeCAFS becomes the strongest method in the study.

This is a defect in the DeCAFS implementation, not a tuning choice: on null
series the older build returned a segmentation whose own penalised cost was
strictly worse than the no-change segmentation under the cost the paper
specifies.

# The SVP within-segment cost

SVP minimises the number of valid segments first and uses a within-segment cost
only to rank the partitions that attain that number, so the cost decides where
the boundaries go and never how many there are. The study passes
`cost = "ar1"`, which scores a segment \((s, t]\) by the sum of squared
deviations of its interior innovations \(Z_u = Y_u - \rho Y_{u-1}\),
\(u = s+2, \dots, t\). The innovation \(Z_{s+1}\) that straddles the
boundary is skipped: its mean is \(\mu_{new} - \rho \mu_{old}\) rather than
\((1-\rho)\mu_{new}\), so charging a segment for it would pull the boundary
away from the change. Every partition into \(K\) segments skips exactly
\(K-1\) innovations, so the comparison SVP actually makes stays fair.

Before this change the study used the Gaussian squared-error cost on the raw
observations, which is misspecified under serial dependence.
`compare_svp_costs.R` runs both costs over the full grid at the same calibrated
threshold and writes `svp_cost_comparison.csv`. Segment counts agree on all
8,400 series; F1 and signal MSE improve in every non-null pattern; and the mean
localisation error given the correct number of changes falls from 1.61 to 0.31
(`up`), 1.79 to 0.38 (`updown`), and 7.36 to 2.21 (`rand1`).

# Calibration

The DeCAFS AR(1) penalty is `beta = c_D log(n)` with
`modelParam = list(sdEta = 0, sdNu = sqrt(1-rho^2), phi = rho)`. The smallest
candidate whose corrected null F1 is at least 0.99 is selected. The selected
empirical DeCAFS target is then used for matching; in the recorded run it is
0.995, while 0.99 is the nominal target. The
approximate AR(1) PELT penalty and AR1Focus threshold are then selected to have
null F1 closest to the selected DeCAFS target. Calibration uses 1,000 null
series and validation uses 1,000 independent null series; both stages use
deterministic seeds and Wilson confidence intervals.

The approximate PELT candidate applies the existing PELT implementation to
the innovations

\[
Z_t = Y_{t+1} - \rho Y_t,
\]

with penalty `c_P log(n)`. A transformed endpoint `j` is mapped back to the
original endpoint `j+1`; the transformed terminal endpoint is discarded and
the original endpoint `n` is appended. This is approximate because it does not
retain the complete boundary correction of the exact AR(1) likelihood.

`PELT inflated` remains fixed at the paper-study value
`3*log(n)*(1+rho)/(1-rho)`. Its null F1 is reported as a fixed-baseline
benchmark and is not used as a calibration target. The approximate PELT
penalty is calibrated alongside DeCAFS and AR1Focus.

```{r calibration-summary, eval=FALSE}
calibration <- run_ar1_calibration(workers = 1L)
calibration$selected
calibration$validation
```

# Methods in the power study

The final four methods are exactly:

1. `PELT AR1 approximate`;
2. `PELT inflated`;
3. `DeCAFS AR1`;
4. `SVP AR1Focus + AR1 cost / right / c = ...`.

The SVP call uses `test = "AR1Focus"`, `subtests = "right"`, and
`cost = "ar1"` only. No multiscale or `both` configuration is part of this
study.

# Reproduction

Run from the package root:

```{r reproduction, eval=FALSE}
calibration <- run_ar1_calibration(workers = 8L)
run_and_save_ar1(workers = 8L, calibration = calibration)
source("simulations/power_ar1/compare_svp_costs.R")
source("simulations/power_ar1/export_paper_figures.R")
export_ar1_paper_figures(calibration = calibration)
```

The seven study PDFs are written to `simulations/power_ar1/plots/`. The
paper-facing Section 5.3 PDFs are written to the sibling `SVP_NEW_Figures2`
directory using the `5_` filename convention. The calibration table records
all candidates, null F1 values, confidence intervals, seeds, selected values,
and validation results.

# Outcome

```{r outcome}
results <- readRDS(file.path("simulations", "power_ar1", "results.rds"))
calibration <- readRDS(
  file.path("simulations", "power_ar1", "ar1_calibration.rds")
)
knitr::kable(
  dplyr::summarise(
    dplyr::group_by(results, algorithm, pattern),
    F1 = mean(F1), Precision = mean(Precision), Recall = mean(Recall),
    CorrectNumCP = mean(CorrectNumCP), MSE = mean(MSE),
    NumSegments = mean(NumSegments), .groups = "drop"
  ),
  digits = 3
)
```

With DeCAFS corrected it is the strongest method on every non-null pattern,
which is consistent with its exact AR(1) penalised-likelihood optimisation,
while approximate PELT drops the dependence across segments and SVP uses a
conditional validity test together with an approximate boundary-excluded cost.
Approximate PELT remains
the weakest of the four. Its transformed series has a spike at every change,
`Z_{tau+1} = Y_{tau+1} - rho Y_tau`, whose mean is
`mu_new - rho mu_old`; PELT fits that spike with short spurious segments, which
is why it detects around ten segments where the truth has eight and why its
localisation error is large even when the segment count happens to be right.
Skipping that innovation is the defining approximation in the SVP AR(1) cost.
